---
title: "Problem Set 03"
author: "Lorenzo Bartolo"
date: "10/18/2021"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE, echo = TRUE)

# List of packages you want to install -- separated with a comma and surrounded in "quotes" 
packages <- c("tidyverse", "dplyr", "moments", "lmtest", "sandwich", "stringr", "readr", "here", "ggplot2", "modelr", "MASS", "knitr", "formatR", "car")

# Installs packages
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Loads packages
invisible(lapply(packages, library, character.only = TRUE))

# Load in data
rural_atlas_merged <- read_csv(here("Data/rural_atlas_merged.csv"))

```


# EMPIRICAL /COMPUTER WORK
5. [50 points (all questions here are worth 5points each except part (a) which is 10 points)] Important: As usual,
your answer should include a printout (can cut and paste into a file. I will show you how to do this) of relevant
calculations on the computer (R or other software output) AND a write up of final answers following the
subparts of the question. The data again are the same as for Problem Set 1. Use these data to answer the
following questions.

(a) It is quite common in econometrics to model income variables nonlinearly. Construct a new variable called
loginc where loginc=ln(PerCapitaInc). Provide summary statistics for this new variable. (Hint: Think back to
how you constructed summary statistics in problem set 1.)
```{r}

```


(b) Now run a model similar to the one that you ran as the final specification in problem set 2 except this time
use loginc as the dependent variable (and we are also going to start controlling for metro status in addition to
the other controls). Specifically, run a regression of loginc with unemployment rate in 2013 (UnempRate2013)
as the main regressor, while also including the other regressors: percentage college educated
(Ed5CollegePlusPct), percentage non-Hispanic black in 2010 (BlackNonHispanicPct2010), percentage
Hispanic in 2010, and metro status in 2013 (Metro2013). Now, what is the estimated effect of UnempRate2013
in words? Also indicate if the relationship is statistically significant at the 10%, 5%, and 1% levels? Make
sure that you are using heteroskedastic standard errors.
```{r}

```


(c) What is the null hypothesis corresponding to the F-statistic as reported in the output for the regression in
part (b)?
```{r}

```


(d) What is the conclusion of the F-test as reported in the output for the regression in part (b)? Explain. (i.e.
Do you reject or fail to reject the null hypothesis above and how do you know this?)
```{r}

```


(e) What are the degrees of freedom (the specific numbers reported) for the F-test as specified in your regression
output? Where are these numbers coming from for this case? (i.e., how is the computer calculating these
numbers specifically given the dataset here?)
```{r}

```


(f) Discuss what the standard error of the regression (SER), R2 and ð‘¹ð‘¹2 in part (b) are telling you in terms of
the numbers that you have found.
```{r}

```


(g) Using what you know about the difference between the two formulas, explain specifically why the R2 and
ð‘¹ð‘¹2 statistics so similar for this case.
```{r}

```


(h) Use an F-test to test the joint significance of the additional regressors: Ed5CollegePlus,
BlackNonHispanicPct2010, HispanicPct2010, and Metro2013. Find this test statistic and clearly indicate the
conclusions of the test.
```{r}

```


(i) If you had more time to study this question and/or more/different data, what would you suggest doing next?
Propose additional variables to add and/or different specifications to try and give specific reasons why you are
suggesting these. Answers will vary for this part of the problem.
```{r}

```


